---
title: "us_higher_ed"
output: html_document
name: Diamond Andy, Lily Gates, Mia Leandri, Colin Thompson
date: 04/10/2025
---
# Overview of Project

## Research Question/Topic
**How do universities compare with the type and amount of research publications?**
  - Academic fields (humanities vs. STEM)
  - Public vs. Private
  - R1 vs. R2
  - Prediction on new policies may affect universities
  - How it may affect admission, funding, research publications etc.
  
## Key Questions:
  - How does the number and type of research publications differ between humanities and STEM fields?
  - What is the relationship between a university’s classification (public/private, R1/R2) and its research output?
  - How might changes in policies regarding funding, research priorities, or admissions affect these relationships?

## Data and Research Sources
**Sources**
*1. Carnegie Classification of Institutions of Higher Education*
  - Classification of R1 vs. R2 Schools
*2. Integrated Postsecondary Education Data System (IPEDS)* 
  - Includes data on U.S. universities
  
**Data Collection**
  - Number and type of publications (e.g., STEM vs. humanities).
  - Institutional classification (Public vs. Private, R1 vs. R2).
  - Funding and admission data.
  - Information on new or proposed policies affecting research funding, publication incentives, or admissions.

-----------------------------------------------
# Importing Libraries
```{r}
library(tidyverse)
library(readr)
library(tidyr)
library(dplyr)
library(lubridate)  # for Datetime formatting
library(stringr)  # for extracting URL
```

# Reading in Data
```{r}
file_name <- read_csv("file_name.csv")
```

# Cleaning and Filtering Data
```{r}

```

# Bar Chart to Compare Publications by Type
  * Across different University classifications
```{r}

```

# Scatter Plot to Examine Relationships Between Funding and Publications
  * Across different University classifications
```{r}

```

# Line Plot to Measure Changes in ___ Overtime
```{r}

```

# Heatmap to Visualize Correlations Between Policy Changes and Outcomes (e.g., admissions, research output, funding)
```{r}

```

